Lossy compression of DNA sequencing quality data

### 2015-02-03 manuscript/q-compression.tex
Lots of changes: new bib entry for Sickle, many text revisions, good/bad is now 2-bin (to emphasize the concept of binning), revised introduction (though it is not yet done).

\section{Results:}
Lossy compression of quality information can greatly decrease storage
and memory requirements with little descernable effects on subsequent
analysis results. The four compression strategies in this study were
able to produce similar results to those obtained with uncompressed
quality sequences in terms of quality control, genome assembly, and
alignment of short read to a reference sequence.

@@

The \emph{de facto} standard for storing next-generation sequencing
data is the \textsc{fastq} format. A component of \textsc{fastq}
format files~\cite{Cock:2010ve} is information for a quantitative
measure of quality for each nucleotide based on the \textsc{phred}
quality value, which is defined as $Q =
-10\ log_{10}\/P$~\cite{Ewing:1998zr,Ewing:1998ly}. Depending on the
sequencing technology these quality values can range from 0 to 93, and
the American Standard Code for Information Interchange \textsc{ascii}
characters 33 to 126 (with some offset) are used to represent quality
values.

Quality values are typically used throughout bioinformatics pipelines.
Certain tools, such as \textsc{fastqc}~\cite{citeulike:11583827} and
Prinseq~\cite{Schmieder:2011gd} provide the user with detailed
descriptions of the distribution of sequence quality scores. These
analyses can then be used to filter and trim the sequencing
data~\cite{EJ200}. The processed sequencing data can then be be used
in subsequent analyses. Certain genome assemblers utilize quality
values directly to produce better
assemblies~\cite{Gnerre:2011kx}. Genome assemblers work by piecing
together overlapping sequences, constructing a larger resulting
sequence. Short-read alignment tools, such as Bowtie
2~\cite{Langmead:2012rw}, use quality values to scale the mismatch
penalty in a way that mismatches between poor quality bases of the
query and the reference are weighted less than mismatches occurring on
high quality bases. Similar to alignment, software for detecting
single nucleotide polymorphisms (\textsc{snp}s) can also rely on
quality values~\cite{McKenna:2010bh}. Detected \textsc{snp}s with
high-quality bases are more trustworthy than those with low-quality
bases, particularly in low coverage regions.

Although the quality values are useful for later steps, the trade-off
between storing effectively double the amount of data and the
resulting effects on subsequent analyses has not been
well-documented. Previous literature on sequence data compression has
largely focused on lossless compression of base
calls~\cite{Giancarlo:2009fk,Nalbantoglu:2010uq,Zhu:2013qr,Deorowicz:2013hq,Giancarlo:2014rw}.
If a reference exists for the biological data, sequences can be
compressed using offsets into the
reference~\cite{Kozanitis:2011kl}. Typical compression methods fail to
utilize the biases inherent in existing sequencing
technologies. Depending on the sequencing technology used, the base
calling is more accurate in the earlier cycles, gradually losing
quality in later cycles~\cite{Kozanitis:2011kl}. To this end, methods
for compressing these quality sequences utilize these general quality
trends. \textsc{SlimGene}~\cite{Kozanitis:2011kl} is one such tool
that builds a fixed-state Markov model on adjacent quality scores and
uses Huffman Coding for compression based on the transition
probabilities.

@@

We examine four general classes of lossy compression methods: score
binning, model-based, profile-based, and data type transformation.

@@

\subsection{Quality value binning}

%The quality value provides some measure of the accuracy of a given
%base call. A strategy to reduce the dimensionality of these quality
%values is to aggregate adjacent quality values together into the same
%bin. Taken to the extreme, a base call can be considered either good
%or bad. This is the motivation of our approach, deemed good/bad
%encoding, where we observe the distribution of quality values across
%the read set and find a cutoff between good and bad bases.

In one of the more extreme cases of quality value binning, we can bin
a quality value into one of two categories: good or bad. We define
good/bad encoding as finding the quality value cutoff to separate
between good and bad bases. First, we learn the distribution of
quality values across all sequences. Using this distribution, bases
are marked bad if their quality value falls below a certain
threshold. We set the cutoff as the first quartile minus 1.5 * the
interquartile range (IQR), which is the difference between the first
and third quartile. 1.5 * IQR is the value used by Tukey's box
plot~\cite{mcgill1978variations}. The main benefit of this approach is
that it is completely data-dependent. No prior assumptions of the
distribution of the quality values need to be made.

@@

Although simple and intuitive, considering a base as either good or
bad may be too limiting for some users, so we provide a few additional
options to preserve more of the defining characteristics of the
quality values, but we only analyze good/bad encoding in our results.

@@

Nonetheless, fitting a polynomial and reconstructing the quality
values that it would have yielded after decompression still makes some
sense as it alters the properties of the quality data, making it
highly compressible simply due to the smoothing effect it
provides. Therefore the compression ratio and bits per bp values for
polynomial regression throughout this manuscript refer to 
polynomially approximated quality values compressed with zip.

@@

K-means clustering is a vector quantization method, partitioning a set
of samples into K sets that minimize within-cluster sum of
squares~\cite{macqueen1967some}. Using a random subset of read quality
values, a compression method can use the computed cluster centers as
read quality profiles. As the problem is NP-hard, we use a heuristic
iterative refinement approach quickly converging to a locally optimal
minimum provided by the Python package scikit-learn
0.15~\cite{pedregosa2011scikit}.

First, the method samples an adjustable number of quality sequences
randomly from the file to be used as a training set. Quality sequences
are represented by vectors containing their \textsc{phred} scores
corresponding to each position along the read. Subsequently, K-means
clustering is performed on the training set until convergence. The
obtained cluster centers will be the quality profile prototypes for
the dataset.

Once the K quality profiles are determined, all quality sequences are
passed through the trained K-means predictor, with the nearest quality
profile in Euclidean space being assigned to every quality sequence as
their compressed representation.

The compressed quality file therefore consists of an index enumerating
the K quality profiles, and a binary part containing the assigned
quality profile index for each read.

@@

\begin{figure}[!tpb]%figure2
\centerline{\includegraphics[width=3.35in]{profiles_128.png}}
\caption{Quality profiles obtained by K-means clustering on the
  fragment library from \textit{Rhodobacter sphaeroides} 2.4.1 data
  set using K=128, with each row corresponding to a quality
  profile. Dark to light colors represent low to high quality
  values. It is readily visible that the two most distinctive features
  of quality profiles is their drop-off position and average overall
  quality. One can also see sporadic low-position values in a handful
  of profiles, likely capturing intermittent problems in the
  sequencing process affecting thousands of reads at a
  time.}\label{fig:profiles_128}
\end{figure}

@@

For our analyses, we use sequencing data provided by the \textsc{GAGE}
project~\cite{Salzberg:2012rc}. The read data for \textit{Rhodobacter
  sphaeroides} 2.4.1 was downloaded from
\href{http://gage.cbcb.umd.edu/data/Rhodobacter_sphaeroides}{http://gage.cbcb.umd.edu/data/Rhodobacter\_sphaeroides},
and the corresponding reference sequence was obtained from the NCBI
RefSeq database (NC\_007488.1, NC\_007489.1, NC\_007490.1,
NC\_007493.1, NC\_007494.1, NC\_009007.1, NC\_009008.1). The
\textit{Rhodobacter sphaeroides} 2.4.1 data set consists of 2,050,868
paired-end reads (insert size of 180 bp) and 2,050,868 short-jump
reads (insert size of 3,000 bp).

@@

For quality control, we use Autoadapt tool which uses
\textsc{fastqc}~\cite{citeulike:11583827} and
\textsc{CutAdapt}~\cite{EJ200} packages with the default
settings. PrinSeq~\cite{Schmieder:2011gd} was used to count the number
of bases and reads in each \textsc{fastq} file.

@@

For aligning the sequences, we use Bowtie 2~\cite{Langmead:2012rw}
with default settings and modified max- and min-mismatch parameters
detailed below.

@@

For any lossy compression method, we need to evaluate the tradeoff
between its physical compression and the loss in its accuracy. Here,
we use average \emph{bits per bp} to measure the physical compression
of the data and mean squared error (MSE) to measure the loss in
quality. Bits per bp is calculated simply as the compressed file size
(in bits) divided by the total number of quality values. MSE is
calculated by $\frac{1}{n}\sum_{i=1}^{n}{(Q_i-Q_i')^2}$, where $n$ is
the number of sequences, $Q_i$ is the actual quality value associated
with sequence position $i$, and $Q_i'$ is the corresponding compressed
quality value.

We compare the \textsc{mse} versus \emph{bits per bp} of the fragment
and short-jump libraries of the \textit{Rhodobacter sphaeroides} 2.4.1
data set (Figures \ref{fig:mse_vs_bpbp_frag} and
\ref{fig:mse_vs_bpbp_jump}). Storing the uncompressed quality values
requires 1 \emph{byte} per basepair because they are stored in
\textsc{ascii} format; however, lossless compression of the values with
BZip2 2.112 bits per basepair. 0-order polynomial regression and
32-profile encodings have the lowest bits per bp. 5th-order polynomial
regression has the highest bits per bp,
but have among the lowest \textsc{mse}. As the order of the polynomial increases, the bits per bp increase and
the \textsc{mse} decreases at an exponential rate.

\begin{figure}[!tpb]%figure2
\centerline{\includegraphics[width=3.65in]{mse_frag.png}}
\caption{MSE versus bits per bp for different compression methods
  applied to the fragment library from \textit{Rhodobacter
    sphaeroides} 2.4.1 data set. G/B: Good/Bad encoding. P32 and P256:
  Quality profile compression with 32 and 256 profiles. R0-R5:
  Polynomial regression-based compression using 0-5 degrees. Dashed line corresponds to lossless
  bzip compression of quality values. \em}\label{fig:mse_vs_bpbp_frag}
\end{figure}

\begin{figure}[!tpb]%figure2
\centerline{\includegraphics[width=3.65in]{mse_short.png}}
\caption{MSE versus bits per bp for for different compression methods
  applied to the short-jump library from \textit{Rhodobacter
    sphaeroides} 2.4.1 data set. G/B: Good/Bad encoding. P32 and P256:
  Quality profile compression with 32 and 256 profiles. R0-R5:
  Polynomial regression-based compression using 0-5 degrees. Dashed line corresponds to
  lossless bzip compression of quality
  values. \em}\label{fig:mse_vs_bpbp_jump}
\end{figure}

@@

One way to evaluate different compression approaches is to perform
quality preprocessing on the generated \textsc{fastq} files and
compare the number of reads and bases with those of the uncompressed
\textsc{fastq} files. Autoadapt is a tool that uses \textsc{fastqc}
and \textsc{Cutadapt} to filter and trim reads. We used PrinSeq to
calculate the number of reads and bases in each sample. Table
\ref{tab:quality_control} shows the proportional numbers with respect
to those of the uncompressed \textsc{fastq} files. In the fragment
sample, we retain on average 95.4\% and 94.24\% of bases and reads,
respectively. In the short-jump sample, we retain 90.4\% and 90.98\%
of bases and reads, respectively. In some cases, the proportion of
reads and/or bases compared to the original is more than 100\%; this
means that the number of the reads and bases in the compressed samples
are more because altering quality values may prevent the read from
being filtered and/or trimmed. This is clear in the good/ bad
compression approach of the short-jump sample, where the proportion of
bases and reads are 113.09\% and 103.70\% respectively.

@@

\textsc{allpaths-lg} is one of the leading bacterial genome
assemblers~\cite{Salzberg:2012rc}. We examine the effect of our lossy
compression methods on the \textit{Rhodobacter sphaeroides} 2.4.1 data
set provided by \textsc{gage} (Genome Assembly Gold-Standard
Evaluations)~\cite{Salzberg:2012rc}. \textsc{allpaths-lg} is used to
assemble the original data along with the quality compressed versions
using good/bad encoding, 0--5th-order polynomial regression encodings,
and 32- and 256-profile encodings. We evaluate
the assembly based on contiguity statistics, log average read
probability (LAP~\cite{Ghodsi:2013hb}), and a collection of
reference-based metrics. The contiguity statistics include number of
scaffolds and NG50, which is defined as the weighted median contig
size (the length of largest contig $c$ such that the total size of the
contigs larger than $c$â exceeds half of the known genome size). The
LAP score can be viewed as a log likelihood score, where a value
closer to 0 is better. We use a reference-based evaluation script
provided by \textsc{gage} which counts single nucleotide polymorphisms
(\textsc{snp}s), relocations, translations, and inversions. The
reference-based metrics are normalized by the length of the assembly
to aid in comparison. Alongside the raw metrics, we also provide a
cummulative sum of the individual ranks of each method for each
metric. A similar approach was utilized by the \textsc{Assemblathon1}
competition~\cite{Earl:2011fv}.

The assembly using the uncompressed reads outperforms all compression
methods in terms of LAP, NG50, relocations, translations, and
inversions (Table \ref{tab:assembly_ranks}). Among the compression
methods, the 
256- and 32-profile encoding had the highest rank, then the 4th-order polynomial
regression, the 5th-order polynomial regression, the good/bad
encoding, and lastly, the 3rd through 0-order polynomials.

The lossy compression methods largely preserve the contiguity found in
the assembly produced using the reads with unmodified quality
sequences. All compression methods other than 0-order polynomial
regression produce an NG50 ranging from 3.18--3.19 Mbps. Despite the
similar contiguity statistics, the different compression methods vary
noticeably in the amount of SNPs. The order of polynomial has an
inverse relationship with the amount of SNPs detected. The good/bad and
profile methods detected the least amount of SNPs compared to
the reference genome, outperforming the assembly using the original
quality sequences. A more in-depth evaluation is needed to determine
whether these compression methods are missing actual SNPs.

It is important to highlight the result that the 4th-order polynomial
regression outperforms the 5th-order in all of the reference-based
metrics, but not the \emph{de novo} LAP metrics. The 5th-order
polynomial regression assembles approximately 2.5 kb more sequence and
uses roughly $1.2\%$ more of the total read set than the 4th-order
(Supplemental Table 1), which may explain the descrepency in LAP
values because the LAP value is influenced by total
coverage~\cite{Ghodsi:2013hb}.

\begin{table*}[!tpb]
\centering
\caption[]{Rankings of different assemblies produced using proposed
  compression methods. Assemblies were constructed using
  \textsc{allpaths-lg} and the \textit{Rhodobacter sphaeroides}
  2.4.1. data set provided by
  \textsc{gage}~\cite{Salzberg:2012rc}. Regression methods are sorted
  by their overall rank (determined by the sum of individual
  rankings).}
{\begin{tabular}{lcccccccc}
& Overall & & & & & & LAP & LAP \\
Compression            & rank & \textsc{snp}s & Inversions & Relocations & Translocations & Indels & fragment & short-jump \\
\hline
None                   & 1    & 5    & 1          & 1           & 1              & 1      & 1          & 1                    \\
Profile (256)          & 3    & 2    & 1          & 3           & 3              & 4      & 2          & 4                    \\
Profile (32)           & 4    & 3    & 1          & 4           & 6              & 5      & 3          & 3                    \\
Regression (4th order) & 5    & 6    & 1          & 6           & 5              & 2      & 7          & 8                    \\
Regression (5th order) & 6    & 7    & 1          & 5           & 7              & 7      & 5          & 7                    \\
Good/bad               & 7    & 1    & 1          & 8           & 11             & 10     & 6          & 5                    \\
Regression (3rd order) & 8    & 8    & 1          & 9           & 9              & 6      & 8          & 6                    \\
Regression (2nd order) & 9    & 9    & 1          & 7           & 4              & 8      & 9          & 10                   \\
Regression (1st order) & 10   & 10   & 11         & 10          & 8              & 9      & 10         & 9                    \\
Regression (0 order) & 11   & 11   & 10         & 11          & 10             & 11     & 11         & 11
\\ \hline
\end{tabular}
}
\label{tab:assembly_ranks}
\end{table*}

@@

Certain short read alignment tools use the quality sequence
information when finding potential alignments. Bowtie 2 (version
2.2.3) was used to evaluate the different decompressed \textsc{fastq}
files. Bowtie 2 uses quality values written in the \textsc{fastq}
files when mapping reads against a reference genome. The original
uncompressed and decompressed \textsc{fastq} files were mapped with
Bowtie 2 against \textit{Rhodobacter sphaeroides} reference
genome. The generated \textsc{sam} file for each compressing approach
were compared with the uncompressed \textsc{sam} file. The total,
shared and unique proportional numbers of mapped reads are calculated
with respect to the uncompressed \textsc{sam} matches as shown in
table \ref{tab:aligner}. Additionally, to monitor the effect of
quality values on mapping in general, Bowtie 2 was tweaked with the
maximum and minimum mismatch penalty equivalent to maximum and minimum
quality scores (with parameters: --mp 6,6 and --mp 2,2 respectively).

@@

	&Maximum quality & 78.08 & 78.08 & 21.92 & 0 \\
	&Minimum quality & 100 & 100 & 0 & 0 \\ 
	&Good/Bad & 97.38 & 97.38 & 2.62 & 0 \\
	&Regression (0 order) & 96.07 & 94.11 & 5.89 & 1.96 \\
	&Regression (1st order) & 99.24 & 98.98 & 1.02 & 0.26 \\ 
	&Regression (2nd order) & 99.66 & 99.35 & 0.65 & 0.3 \\ 
	Short-jump&Regression (3rd order) & 99.94 & 99.61 & 0.39 & 0.33 \\ 
	&Regression (4th order) & 99.99 & 99.64 & 0.36 & 0.35 \\ 
	&Regression (5th order) & 100 & 99.67 & 0.33 & 0.33 \\ 
	&Profile (32) & 99.74 & 99.44 & 0.56 & 0.3 \\
	&Profile (256) & 99.76 & 99.52 & 0.48 & 0.24 \\
\end{tabular}

@@

\subsection{Extension of good/bad encoding}

Good/bad encoding has the nice property of being simple to compute and
has a good bits per bp score. The good/bad encoding suffers from
having a high \textsc{mse}, but fortunately, we have shown that in the
case of genome assembly, good/bad encoding outperforms all polynomial
regressions encodings with degree less than 3. Good/bad encoding of
the fragment and short-jump libraries of \textit{Rhodobacter
  sphaeroides} have \textsc{mse}s of $2.42\times$ and $10.76\times$
the 3rd-order polynomial regression encodings, respectively. This
further highlights the importance of using additional contextual
information of the subsequent analyses when evaluating compressed
quality values.

>>>>

\section{Results:}
Lossy compression of quality information can greatly decrease storage
and memory requirements with little discernable effects on subsequent
analysis results. The four compression strategies in this study were
able to produce similar results to those obtained with uncompressed
quality sequences in terms of quality control, genome assembly, and
alignment of short read to a reference sequence.

@@

Sequencing read data constitutes the largest category of sequence data
in genomics research because of great redundancy, inclusion of quality
values, and read-level naming and metadata. Because of this abundance
effective compression of read data has the potential for substantial
improvement in data storage and transfer efficiency.  Among the
several challenges for compression of read data is dealing with
different error profiles resulting from differences in underlying
chemistries, signal detection and processing mechanisms, inherent
biases, and other idiosyncratic properties of distinct high-throughput
sequencing technologies. Illumina sequencing reads, the focus of this
research, are characterized by having relatively few insertion and
deletion errors, but substitution (miscall) errors are much more
frequent and have context-specific patterns. These errors are
non-uniformly distributed over the read length (e.g., error rates
increase up to $\sim$16$\times$ at the 3$^{\prime}$ end, and 32.8 --
67.9\% of reads have low quality tails at the 3$^{\prime}$
end~\cite{Minoche:2011km}).

Quality values comprise a standard component of \textsc{fastq} format
files~\cite{Cock:2010ve}.  At the level of sequence read the
probability of error for each base-call is typically represented by
\textsc{phred} quality value, which is defined as $Q =
-10\,log_{10}P$~\cite{Ewing:1998ly}. Depending on the sequencing
technology these quality values can range from 0 to 93, and are
represented with the \textsc{ascii} characters 33 to 126 (with some
offset). These errors are represented a single quality value per
base-call for Illumina sequencing reads.

Quality values are typically used throughout bioinformatics pipelines.
Software such as \textsc{fastqc}~\cite{citeulike:11583827} and
Prinseq~\cite{Schmieder:2011gd} provide the user with detailed
descriptions of the distribution of sequence quality scores, and these
scores can be used to guide filtering and trimming of sequencing read
data as processing step before subsequent analyses. Quality values can
be used by genome assembly software to produce better
assemblies~\cite[e.g.,][]{Gnerre:2011kx}.  Short-read alignment
software, such as Bowtie 2~\cite{Langmead:2012rw}, use quality values
to scale the mismatch penalty in a way that mismatches between poor
quality bases of the query and the reference are weighted less than
mismatches occurring on high quality bases. Similar to alignment,
software for detecting single nucleotide polymorphisms (\textsc{snp}s)
can use quality values~\cite[e.g.,][]{McKenna:2010bh}, and identified
\textsc{snp}s with high-quality calls are deemed more reliable than
those with low-quality calls, particularly in low coverage regions.

Although the quality values are useful for later steps, the trade-off
between storing effectively double the amount of data and the
resulting effects on subsequent analyses has not been
well-documented. Previous literature on sequence data compression has
largely focused on lossless compression of base
calls~\cite{Giancarlo:2009fk,Nalbantoglu:2010uq,Zhu:2013qr,Deorowicz:2013hq,Giancarlo:2014rw}.
Typical compression methods fail to utilize the biases inherent in
existing sequencing technologies. Depending on the sequencing
technology used, the base calling is more accurate in the earlier
cycles, gradually losing quality in later
cycles~\cite{Kozanitis:2011kl}. To this end, methods for compressing
these quality sequences utilize these general quality
trends. \textsc{SlimGene}~\cite{Kozanitis:2011kl} is one such tool
that builds a fixed-state Markov model on adjacent quality scores and
uses Huffman Coding for compression based on the transition
probabilities.

@@

We examine three general classes of lossy compression methods:
binning, model-based, and profile-based.

@@

\subsection{Binning}

Quality values can be binned into a set of bins, and any number of
bins less than the the number of distinct quality values is
possible. The minimum number of bins that allows for a any distinction
among quality values is two, i.e., two categories ``good'' and ``bad''
quality. We implement 2-bin encoding by setting a quality value
threshold empirically determined by the distribution of quality values
across reads. Base-calls are marked ``bad'' if their quality value
falls below the first quartile minus 1.5 $\times$ the interquartile
range (IQR), which is the difference between the first and third
quartile. 1.5 $\times$ IQR is the value used by Tukey's box
plot~\cite{mcgill1978variations}. The main benefit of this approach is
that it is completely data-dependent. No prior assumptions of the
distribution of the quality values need to be made.

@@

Although simple and intuitive, considering a base as either good or
bad may be too limiting for some users, so we provide a few additional
options to preserve more of the defining characteristics of the
quality values, but we only analyze 2-bin encoding in our results.

@@

Nonetheless, fitting a polynomial and reconstructing the quality
values that it would have yielded after decompression still makes some
sense as it alters the properties of the quality data, making it
highly compressible simply due to the smoothing effect it
provides. Therefore the compression ratio and bits/base-call values
for polynomial regression throughout this manuscript refer to
polynomially approximated quality values compressed with zip.

@@

$k$-means clustering is a vector quantization method, partitioning a set
of samples into $k$ sets that minimize within-cluster sum of
squares~\cite{macqueen1967some}. Using a random subset of read quality
values, a compression method can use the computed cluster centers as
read quality profiles. As the problem is NP-hard, we use a heuristic
iterative refinement approach quickly converging to a locally optimal
minimum provided by the Python package scikit-learn
0.15~\cite{pedregosa2011scikit}.

First, the method samples an adjustable number of quality sequences
randomly from the file to be used as a training set. Quality sequences
are represented by vectors containing their \textsc{phred} scores
corresponding to each position along the read. Subsequently, $k$-means
clustering is performed on the training set until convergence. The
obtained cluster centers will be the quality profile prototypes for
the dataset.

Once the $k$ quality profiles are determined, all quality sequences are
passed through the trained $k$-means predictor, with the nearest quality
profile in Euclidean space being assigned to every quality sequence as
their compressed representation.

The compressed quality file therefore consists of an index enumerating
the $k$ quality profiles, and a binary part containing the assigned
quality profile index for each read.

@@

\begin{figure}[!tpb]%figure2
\centerline{\includegraphics[width=3.35in]{profiles_128.png}}
\caption{Quality profiles obtained by $k$-means clustering on the
  fragment library from \textit{Rhodobacter sphaeroides} 2.4.1 data
  set using $k$ = 128, with each row corresponding to a quality
  profile. Dark to light colors represent low to high quality
  values. It is readily visible that the two most distinctive features
  of quality profiles is their drop-off position and average overall
  quality. One can also see sporadic low-position values in a handful
  of profiles, likely capturing intermittent problems in the
  sequencing process affecting thousands of reads at a
  time.}\label{fig:profiles_128}
\end{figure}

@@

We used several Illumina sequence read data sets in this research as
follows.

\textit{Rhodobacter sphaeroides} 2.4.1 data from the \textsc{gage}
(Genome Assembly Gold-Standard Evaluations)~\cite{Salzberg:2012rc},
which are generated from a fragment library (insert size of 180 nt;
2,050,868 paired-end reads) and short-jump library (insert size of
3,000 nt; 2,050,868 reads). The corresponding reference sequence was
obtained from the NCBI RefSeq database (NC\_007488.1, NC\_007489.1,
NC\_007490.1, NC\_007493.1, NC\_007494.1, NC\_009007.1, NC\_009008.1).

\textit{Homo sapiens} \ldots

\textit{Genus speices} \ldots

@@

For quality control, we use
Sickle~\cite{sickle}. PrinSeq~\cite{Schmieder:2011gd} was used to
count the number of bases and reads in each \textsc{fastq} file.

@@

For aligning the sequences, we use Bowtie2~\cite{Langmead:2012rw} with
default settings and modified max- and min-mismatch parameters
detailed below.

@@

For any lossy compression method, we need to evaluate the tradeoff
between its physical compression and the loss in its accuracy. Here,
we use mean bits/base-call to measure the physical compression of the
data and mean squared error (\textsc{mse}) to measure the loss in
quality. Bits/base-call is calculated simply as the compressed file
size (in bits) divided by the total number of quality
values. \textsc{mse} is calculated as
$\frac{1}{n}\sum_{i=1}^{n}{(Q_i-Q_i')^2}$, where $n$ is the number of
sequences, $Q_i$ is the original quality value associated with
sequence position $i$, and $Q_i'$ is the corresponding compressed
quality value.

We compare the \textsc{mse} versus bits/base-call of the fragment and
short-jump libraries of the \textit{Rhodobacter sphaeroides} 2.4.1
data set (Figures \ref{fig:mse_vs_bpbp_frag} and
\ref{fig:mse_vs_bpbp_jump}). Storing the uncompressed quality values
requires 1 byte per base-call because they are stored in
\textsc{ascii} format; however, lossless compression of the values
with BZip2 2.112 bits/base-call. 0-order polynomial regression and
32-profile encodings have the lowest bits/base-call. 5th-order
polynomial regression has the highest bits/base-call, but have among
the lowest \textsc{mse}. As the order of the polynomial increases, the
bits/base-call increase and the \textsc{mse} decreases at an
exponential rate.

\begin{figure}[!tpb]%figure2
\centerline{\includegraphics[width=3.65in]{mse_frag.png}}
\caption{Mean squared error versus bits/base-call for different
  compression methods applied to the fragment library from
  \textit{Rhodobacter sphaeroides} 2.4.1 data set. G/B: 2-bin
  encoding. P32 and P256: Quality profile compression with 32 and 256
  profiles. R0-R5: Polynomial regression-based compression using 0-5
  degrees. Dashed line corresponds to lossless bzip compression of
  quality values. \em}\label{fig:mse_vs_bpbp_frag}
\end{figure}

\begin{figure}[!tpb]%figure2
\centerline{\includegraphics[width=3.65in]{mse_short.png}}
\caption{Mean squared error versus bits/base-call for for different
  compression methods applied to the short-jump library from
  \textit{Rhodobacter sphaeroides} 2.4.1 data set. G/B: 2-bin
  encoding. P32 and P256: Quality profile compression with 32 and 256
  profiles. R0-R5: Polynomial regression-based compression using 0-5
  degrees. Dashed line corresponds to lossless bzip compression of
  quality values. \em}\label{fig:mse_vs_bpbp_jump}
\end{figure}

@@

One way to evaluate different compression approaches is to perform
quality preprocessing on the generated \textsc{fastq} files and
compare the number of reads and bases with those of the uncompressed
\textsc{fastq} files. Sickle is a tool to filter and trim reads. We
used PrinSeq to calculate the number of reads and bases in each
sample. Table \ref{tab:quality_control} shows the proportional numbers
with respect to those of the uncompressed \textsc{fastq} files. In the
fragment sample, we retain on average 95.4\% and 94.24\% of bases and
reads, respectively. In the short-jump sample, we retain 90.4\% and
90.98\% of bases and reads, respectively. In some cases, the
proportion of reads and/or bases compared to the original is more than
100\%; this means that the number of the reads and bases in the
compressed samples are more because altering quality values may
prevent the read from being filtered and/or trimmed. This is clear in
the good/ bad compression approach of the short-jump sample, where the
proportion of bases and reads are 113.09\% and 103.70\% respectively.

@@

\textsc{allpaths-lg} is one of the leading bacterial genome
assemblers~\cite{Salzberg:2012rc}. We examine the effect of our lossy
compression methods on the \textit{Rhodobacter sphaeroides} 2.4.1 data
set provided by \textsc{gage} (Genome Assembly Gold-Standard
Evaluations)~\cite{Salzberg:2012rc}. \textsc{allpaths-lg} is used to
assemble the original data along with the quality compressed versions
using 2-bin encoding, 0--5th-order polynomial regression encodings,
and 32- and 256-profile encodings. We evaluate the assembly based on
contiguity statistics, log average read probability
(\textsc{lap}~\cite{Ghodsi:2013hb}), and a collection of
reference-based metrics. The contiguity statistics include number of
scaffolds and NG50, which is defined as the weighted median contig
size (the length of largest contig $c$ such that the total size of the
contigs larger than $c$â exceeds half of the known genome size). The
\textsc{lap} score can be viewed as a log likelihood score, where a
value closer to 0 is better. We use a reference-based evaluation
script provided by \textsc{gage} which counts single nucleotide
polymorphisms (\textsc{snp}s), relocations, translations, and
inversions. The reference-based metrics are normalized by the length
of the assembly to aid in comparison. Alongside the raw metrics, we
also provide a cummulative sum of the individual ranks of each method
for each metric. A similar approach was utilized by the
\textsc{Assemblathon1} competition~\cite{Earl:2011fv}.

The assembly using the uncompressed reads outperforms all compression
methods in terms of \textsc{lap}, NG50, relocations, translations, and
inversions (Table \ref{tab:assembly_ranks}). Among the compression
methods, the 256- and 32-profile encoding had the highest rank, then
the 4th-order polynomial regression, the 5th-order polynomial
regression, the 2-bin encoding, and lastly, the 3rd through 0-order
polynomials.

The lossy compression methods largely preserve the contiguity found in
the assembly produced using the reads with unmodified quality
sequences. All compression methods other than 0-order polynomial
regression produce an NG50 ranging from 3.18--3.19 Mbps. Despite the
similar contiguity statistics, the different compression methods vary
noticeably in the amount of \textsc{snp}s. The order of polynomial has
an inverse relationship with the amount of \textsc{snp}s detected. The
2-bin and profile methods detected the least amount of \textsc{snp}s
compared to the reference genome, outperforming the assembly using the
original quality sequences. A more in-depth evaluation is needed to
determine whether these compression methods are missing actual
\textsc{snp}s.

It is important to highlight the result that the 4th-order polynomial
regression outperforms the 5th-order in all of the reference-based
metrics, but not the \emph{de novo} \textsc{lap} metrics. The
5th-order polynomial regression assembles approximately 2.5 kb more
sequence and uses roughly $1.2\%$ more of the total read set than the
4th-order (Supplemental Table 1), which may explain the discrepency in
\textsc{lap} values because the \textsc{lap} value is influenced by
total coverage~\cite{Ghodsi:2013hb}.

\begin{table*}[!tpb]
\centering
\caption[]{Rankings of different assemblies produced using proposed
  compression methods. Assemblies were constructed using
  \textsc{allpaths-lg} and the \textit{Rhodobacter sphaeroides}
  2.4.1. data set provided by
  \textsc{gage}~\cite{Salzberg:2012rc}. Regression methods are sorted
  by their overall rank (determined by the sum of individual
  rankings).}
{\begin{tabular}{lcccccccc}
& Overall & & & & & & \textsc{lap} & \textsc{lap} \\
Compression            & rank & \textsc{snp}s & Inversions & Relocations & Translocations & Indels & fragment & short-jump \\
\hline
None                   & 1    & 5    & 1          & 1           & 1              & 1      & 1          & 1                    \\
Profile (256)          & 3    & 2    & 1          & 3           & 3              & 4      & 2          & 4                    \\
Profile (32)           & 4    & 3    & 1          & 4           & 6              & 5      & 3          & 3                    \\
Regression (4th order) & 5    & 6    & 1          & 6           & 5              & 2      & 7          & 8                    \\
Regression (5th order) & 6    & 7    & 1          & 5           & 7              & 7      & 5          & 7                    \\
2-bin	               & 7    & 1    & 1          & 8           & 11             & 10     & 6          & 5                    \\
Regression (3rd order) & 8    & 8    & 1          & 9           & 9              & 6      & 8          & 6                    \\
Regression (2nd order) & 9    & 9    & 1          & 7           & 4              & 8      & 9          & 10                   \\
Regression (1st order) & 10   & 10   & 11         & 10          & 8              & 9      & 10         & 9                    \\
Regression (0 order) & 11   & 11   & 10         & 11          & 10             & 11     & 11         & 11
\\ \hline
\end{tabular}
}
\label{tab:assembly_ranks}
\end{table*}

@@

Certain short read alignment tools use the quality sequence
information when finding potential alignments. Bowtie2 (version 2.2.3)
was used to evaluate the different decompressed \textsc{fastq}
files. Bowtie2 uses quality values written in the \textsc{fastq} files
when mapping reads against a reference genome. The original
uncompressed and decompressed \textsc{fastq} files were mapped with
Bowtie2 against \textit{Rhodobacter sphaeroides} reference genome. The
generated \textsc{sam} file for each compressing approach were
compared with the uncompressed \textsc{sam} file. The total, shared
and unique proportional numbers of mapped reads are calculated with
respect to the uncompressed \textsc{sam} matches as shown in table
\ref{tab:aligner}. Additionally, to monitor the effect of quality
values on mapping in general, Bowtie2 was tweaked with the maximum and
minimum mismatch penalty equivalent to maximum and minimum quality
scores (with parameters: --mp 6,6 and --mp 2,2 respectively).

@@

	&Maximum quality & 78.08 & 78.08 & 21.92 & 0 \\
	&Minimum quality & 100 & 100 & 0 & 0 \\ 
	&2-bin & 97.38 & 97.38 & 2.62 & 0 \\
	&Regression (0 order) & 96.07 & 94.11 & 5.89 & 1.96 \\
	&Regression (1st order) & 99.24 & 98.98 & 1.02 & 0.26 \\ 
	&Regression (2nd order) & 99.66 & 99.35 & 0.65 & 0.3 \\ 
	Short-jump&Regression (3rd order) & 99.94 & 99.61 & 0.39 & 0.33 \\ 
	&Regression (4th order) & 99.99 & 99.64 & 0.36 & 0.35 \\ 
	&Regression (5th order) & 100 & 99.67 & 0.33 & 0.33 \\ 
	&Profile (32) & 99.74 & 99.44 & 0.56 & 0.3 \\
	&Profile (256) & 99.76 & 99.52 & 0.48 & 0.24 \\
\end{tabular}

@@

\subsection{Extension of 2-bin encoding}

2-bin encoding has the nice property of being simple to compute and
has good bits/base-call values. The 2-bin encoding suffers from having
a high \textsc{mse}, but fortunately, we have shown that in the case
of genome assembly, 2-bin encoding outperforms all polynomial
regressions encodings with degree less than 3. 2-bin encoding of the
fragment and short-jump libraries of \textit{Rhodobacter sphaeroides}
have \textsc{mse}s of $2.42\times$ and $10.76\times$ the 3rd-order
polynomial regression encodings, respectively. This further highlights
the importance of using additional contextual information of the
subsequent analyses when evaluating compressed quality values.
